Type theory in type theory with single substitutions

### 2024-09-14 types_2024_postproc/p.tex
stuff

\newcommand{\ra}{\rightarrow}
\newcommand{\Set}{\mathsf{Set}}
\newcommand{\Ty}{\mathsf{Ty}}
\newcommand{\Tm}{\mathsf{Tm}}
\newcommand{\Con}{\mathsf{Con}}
\newcommand{\Sub}{\mathsf{Sub}}
\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}
\newcommand{\ext}{\mathop{\triangleright}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\lam}{\mathsf{lam}}
\newcommand{\app}{\mathsf{app}}
\newcommand{\U}{\mathsf{U}}
\newcommand{\El}{\mathsf{El}}
\newcommand{\cd}{\mathsf{c}}
\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}} %from the book

@@

The single substitution semantics is a GAT which does not start with a
category: there are contexts and maps between contexts, but these maps
are single substitutions or single weakenings, they don't
compose. There are fewer operations and fewer equations than in the
parallel semantics. In this paper we investigate the single
substitution semantics of the type theory specified by the SOGAT
(\ref{eq:tt}). This presentation of type theory is simpler than CwFs,
the operations are all easy to motivate, we illustrate this in Section
\ref{sec:ssc}. Single substitution calculi are closer to the usual
presentation of type theory using derivation rules, this was one of
the motivations of Voevodsky for developing B-systems
\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}. Our single substitution
semantics of (\ref{eq:tytm}) is simpler and has more models than
B-systems. GAT presentations of type theory are important for defining
models, our presentation gives a new notion of model, which in the
future might contribute to definitions of new models. Because our
single substitution GAT does not involve a category, it might be
easier to adapt it in a higher dimensional setting, see Section
\ref{sec:conclusion}.

@@

if we have U,Pi, we can minimise the syntax

if we have U,Pi,Sigma, then SSC semantics of (1)+Sigma -> PSC semantics of (1)+Sigma.

@@

Contents of the TYPES abstract, but with more details.
\begin{alignat*}{10}
& \Con && : \Set \\
& \Ty && : \Con\ra\N\ra\Set \\
& \Tm && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Set \\
& \diamond && : \Con \\
& \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
& \Sub && : \Con\ra\Con\ra\Set \\
& \blank[\blank] && : \Ty\,\Gamma\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta \\
& \blank[\blank] && : \Tm\,\Gamma\,A\ra(\gamma:\Sub\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\gamma]) \\
& \p && : \Sub\,(\Gamma\ext A)\,\Gamma \\
& \q && : \Tm\,(\Gamma\ext A)\,(A[\p]) \\
& \Pi && : (A:\Ty\,\Gamma)\ra\Ty\,(\Gamma\ext A)\ra\Ty\,\Gamma \\
& \blank^+ && : (\gamma:\Sub\,\Delta\,\Gamma)\ra\Sub\,(\Delta\ext A[\gamma])\,(\Gamma\ext A) \\
& \Pi[] && : (\Pi\,A\,B)[\gamma] = \Pi\,(A[\gamma])\,(B[\gamma^+]) \\
& \lam && : \Tm\,(\Gamma\ext A)\,B\ra\Tm\,\Gamma\,(\Pi\,A\,B) \\
& \lam[] && : (\lam\,t)[\gamma] = \lam\,(t[\gamma^+]) \\
& \langle\blank\rangle && : \Tm\,\Gamma\,A\ra\Sub\,\Gamma\,(\Gamma\ext A) \\
& \blank\cdot\blank && : \Tm\,\Gamma\,(\Pi\,A\,B)\ra(a:\Tm\,\Gamma\,A)\ra\Tm\,\Gamma\,(B[\langle a\rangle]) \\
& \Pi\beta && : \lam\,t\cdot a = t[\langle a\rangle] \\
& [\langle\rangle] && : B[\langle a\rangle][\gamma] = B[\gamma^+][\langle a[\gamma]\rangle] \\
& {\cdot}[] && : (t\cdot a)[\gamma] = (t[\gamma])\cdot(a[\gamma]) \\
& [\p][^+] && : B[\p][\gamma^+] = B[\gamma][\p] \\
& [\p][^+] && : b[\p][\gamma^+] = b[\gamma][\p] \\
& \q[^+] && : \q[\gamma^+] = \q \\
& [\p][\langle\rangle] && : B[\p][\langle a\rangle] = B \\
& [\p][\langle\rangle] && : b[\p ][\langle a\rangle] = b \\
& \q[\langle\rangle] && : \q[\langle a\rangle] = a \\
& [\p^+][\langle\q\rangle] && : B[\p^+][\langle\q\rangle] = B \\
& \Pi\eta && : t = \lam\,(t[\p]\cdot\q) \\
\end{alignat*}

@@

\section{Isomorphism}
\label{sec:iso}

>>>>

\newcommand{\ra}{\rightarrow}
\newcommand{\Ra}{\Rightarrow}
\newcommand{\Set}{\mathsf{Set}}
\newcommand{\Ty}{\mathsf{Ty}}
\newcommand{\Tm}{\mathsf{Tm}}
\newcommand{\Con}{\mathsf{Con}}
\newcommand{\Sub}{\mathsf{Sub}}
\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}
\newcommand{\ext}{\mathop{\triangleright}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\lam}{\mathsf{lam}}
\newcommand{\app}{\mathsf{app}}
\newcommand{\U}{\mathsf{U}}
\newcommand{\El}{\mathsf{El}}
\newcommand{\cd}{\mathsf{c}}
\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}} %from the book
\renewcommand{\tt}{\mathsf{tt}}
\newcommand{\fst}{\mathsf{fst}}
\newcommand{\snd}{\mathsf{snd}}

@@

The single substitution semantics is a GAT which does not start with a
category: there are contexts and maps between contexts, but these maps
are single substitutions or single weakenings, they don't
compose. There are fewer operations and fewer equations than in the
parallel semantics. In this paper we investigate the single
substitution semantics of the type theory specified by the SOGAT
(\ref{eq:tt}). This presentation of type theory is simpler than CwFs,
the operations are all easy to motivate, we illustrate this in Section
\ref{sec:ssc}. Single substitution calculi are closer to the usual
presentation of type theory using derivation rules, this was one of
the motivations of Voevodsky for developing B-systems
\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}. Our single substitution
semantics of (\ref{eq:tytm}) is simpler and has more models than
B-systems. GAT presentations of type theory are important for defining
models, our presentation gives a new notion of model, which in the
future might contribute to definitions of new models. It can also
influence implementations of type theory. Because our single
substitution GAT does not involve a category, it might be easier to
adapt it to a higher dimensional setting, see Section
\ref{sec:conclusion}.

@@

In the particular theory (\ref{eq:tt}), every type is represented by a
term. In the first order single substitution calculus, this can be
used to adapt some equations and remove some others, obtaining a more
minimalistic, but equivalent GAT. Theory (\ref{eq:tt}) can be extended
with unit and $\Sigma$ types, more precisely with the following rules.
\begin{equation}\label{eq:sigma}
  \begin{alignedat}{10}
    & \top && : \Ty\,i \\
    & \tt && : \Tm\,\top \\
    & \top\eta && : t = \tt \\
    & \Sigma && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
    & \fst,\snd && : \Tm\,(\Sigma\,A\,B)\cong(a:\Tm\,A)\times\,\Tm\,(B\,a): \blank,\blank
\end{alignedat}
\end{equation}
Using the same insight about types and the universe, the fact that
dependences can be represented by function spaces, and that
collections of dependencies can be collected using $\Sigma$ types, we
can actually derive a parallel model from a single substitution model.
TODO: universe levels are a problem (maybe not if we use iterated
function space instead of $\Sigma$); the resulting CwF is not the same as
the free CwF over the SSC. Bootstrapping a CwF.
CwF -> SSC -> CwF = termification, that is equationally complete, see shallow paper, so we get an isomorphic CwF
SSC -> CwF -> SSC = we get a stronger thing than what we started with.

@@

In this section, we introduce the syntax of type theory with function
space and universes in a minimalistic way, only introducing operations
that are unavoidable. We eschew boilerplate by only defining
well-formed, well-scoped, well-typed terms that are quotiented by
conversion. This means that the equalities that hold between terms are
the ways we can convert terms to each other when running them as
programs. We give a tutorial-style introduction to the syntax, we do
not assume prior knowledge of the metatheory of type theory. The
definition of type theory that we obtain is the same as the single
substitution GAT semantics \cite{DBLP:conf/fscd/KaposiX24} of the
SOGAT (\ref{eq:tt}).

First of all, we need a sort of terms which have to be indexed by
types because we only want well-typed terms. Both types and terms can
include variables, and to keep track of the currently available
variables, we also index them by contexts: a context is a list of
types, the length is the number of available variables, and we add new
variables at the end (that is, they are snoc-lists and not
cons-lists). Types are also indexed by their universe level, this is a
technical requirement for avoiding Russell's paradox. The index $i$ is
an implicit argument of $\Tm$.
\begin{alignat*}{10}
& \Con && : \Set \\
& \Ty && : \Con\ra\N\ra\Set \\
& \Tm && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Set
\end{alignat*}
Just as lists have two constructors, there are two ways to form
contexts: the empty context $\diamond$ and context extension $\ext$
which is like the snoc operation for lists. Context extension takes a
type which can have variables in the context preceeding the type.
\begin{alignat*}{10}
& \diamond && : \Con \\
& \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Con
\end{alignat*}
We will refer to variables by their distance from the end of the
context. We would like to describe the operation providing the last
variable in a context, the zero De Bruijn index, we denote it by $\q :
\Tm\,(\Gamma\ext A)\,A'$. It is a term in an extended context, and it
should have type $A$, but the issue is that $A : \Ty\,\Gamma$, and $A'
: \Ty\,(\Gamma\ext A)$. We need a way to weaken the type $A$ to obtain
such an $A'$. For this, we will introduce an operation $\blank[\p] :
\Ty\,\Gamma\,i\ra\Ty\,(\Gamma\ext A)\,i$ and define $A' := A[\p]$. Instead
of introducing just $\blank[\p]$, we generalise a bit and add a new
sort $\Sub$ which for now only has the single element $\p$ and the
operation $\blank[\blank]$ is called \emph{instantiation}. We
instantiate the type with the weakening in
$\Sub\,\Delta\,\Gamma$. Elements of $\Sub$ will be in general called
substitutions, hence the name, but for now, we only have single end-of
context weakenings in $\Sub$. $\Delta$ is called the domain, $\Gamma$
the codomain of a $\Sub\,\Delta\,\Gamma$.
\begin{alignat*}{10}
& \Sub && : \Con\ra\Con\ra\Set \\
& \p && : \Sub\,(\Gamma\ext A)\,\Gamma \\
& \blank[\blank] && : \Ty\,\Gamma\,i\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta\,i \\
& \q && : \Tm\,(\Gamma\ext A)\,(A[\p])
\end{alignat*}
The operations $\p$ and $\q$ take three implicit parameters, $\Gamma$,
$i$ and $A$, while $\blank[\blank]$ takes $\Gamma$, $i$ and $\Delta$
implicitly. Still, we only have the last variable $\q$ in the context,
we don't have e.g.\ the last but one variable in $\Tm\,(\Gamma\ext
A\ext B)\,(A[\p][\p])$ where we had to weaken $A$ twice to make it fit
with its context. To obtain more variables, we also allow weakening of
terms, more precisely, we introduce an instantiation operation for
terms with an overloaded notation.
\begin{alignat*}{10}
& \blank[\blank] && : \Tm\,\Gamma\,A\ra(\gamma:\Sub\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\gamma])
\end{alignat*}
Note that this is a dependent function as the type of the resulting
term has to be weakened the same way as the term itself. Now we can
define all variables counting from the end of the context (De Bruijn
indices \cite{debruijn}): $0 := \q$, $1 := \q[\p]$, $2 := \q[\p][\p]$,
$3 := \q[\p][\p][\p]$, and so on.

Next we add dependent function space $\Pi$: the domain of a dependent
function is a type in some context $\Gamma$, and the codomain can also
refer to a variable in the domain, so we extend the context of the
codomain type with the type of the domain. For simplicity, both types
are at the same level.
\begin{alignat*}{10}
& \Pi && : (A:\Ty\,\Gamma\,i)\ra\Ty\,(\Gamma\ext A)\,i\ra\Ty\,\Gamma\,i
\end{alignat*}
The nondependent function space $\blank\Ra\blank :
\Ty\,\Gamma\,i\ra\Ty\,\Gamma\,i\ra\Ty\,\Gamma\,i$ is a special case of $\Pi$
and we define it as the abbreviation $A\Ra B := \Pi\,A\,(B[\p])$.

Because we introduced weakening, we now have to explain what happens
to a $\Pi$ type once we weaken it. Assume $A : \Ty\,\Gamma\,i$,
$B:\Ty\,(\Gamma\ext A)\,i$ and we have $\p:\Sub\,(\Gamma\ext
C)\,\Gamma$. Then we say $(\Pi\,A\,B)[\p] = \Pi\,(A[\p])\,(B[\p'])$,
but $B$ cannot be weakened by $\p$ because there we need $\p' :
\Sub\,(\Gamma\ext C\ext A[\p])\,(\Gamma\ext A)$. So $p'$ has to be a
new kind of weakening which adds a new variable in the last but one
position. But then we need an equation also for $(\Pi\,A\,B)[\p']$,
introducing a new weakening which adds a new variable in the last but
two position, and so on. We solve all of these issues by allowing the
\emph{lifting} of a weakening: $\gamma^+$ will be the same as the
weakening $\gamma : \Sub\,\Delta\,\Gamma$, except that it will add an
extra variable both at the end of the domain and codomain
context.
\begin{alignat*}{10}
& \blank^+ && : (\gamma:\Sub\,\Delta\,\Gamma)\ra\Sub\,(\Delta\ext A[\gamma])\,(\Gamma\ext A)
\end{alignat*}
Note that $A$ is an implicit argument of $\blank^+$, and that it has
to be instantiated by the weakening $\gamma$ in the domain context in
order to fit in. Now we can explain how instantiation acts on $\Pi$ by
the following equation.
\begin{alignat*}{10}
& \Pi[] && : (\Pi\,A\,B)[\gamma] = \Pi\,(A[\gamma])\,(B[\gamma^+])
\end{alignat*}
The above equation has 6 implicit arguments, namely $\Gamma$, $i$,
$A$, $B$, $\Delta$ and $\gamma$.  This rule works for any weakening,
no matter how many $\blank^+$s have been applied to it. As of now, all
elements of $\Sub$ have the form
\begin{alignat*}{10}
  & \p^{+^n} : \Sub\, && (\Gamma\ext A\ext B_1[\p]\ext B_2[\p^+]\ext\dots\ext B_n[\p^{+^{n-1}}]) \\
  & && (\Gamma\ext B_1\ext B_2\ext\dots\ext B_n).
\end{alignat*}
where $+^n$ means the $n$-times iteration of $\blank^+$.

Now that we have weakenings of the form $\gamma^+$, we have to say how
they act on variables, that is, terms of the form
$\q[\p]\dots[\p]$. We express this using two rules: we say what
$\q[\gamma^+]$ computes to, and what $b[\p][\gamma^+]$ computes to
where $b$ is an arbitrary term. $\q[\gamma^+]$ should be the same as
$\q$ (with different implicit arguments as the $\q$ in
$\q[\gamma^+]$), as the weakening happens somewhere in the middle of
the context, so the index of the variable remains unchanged. However,
assuming $\q:\Tm\,(\Gamma\ext A)\,(A[\p])$, we have $\q[\gamma^+] :
\Tm\,(\Delta\ext A[\gamma])\,(A[\p][\gamma^+])$, but in the same
context, we have $\q : \Tm\,(\Delta\ext A[\gamma])\,(A[\gamma][\p])$,
hence the terms in the two sides of the equation $\q[\gamma^+] = \q$
have different types. But these two types should be the same:
weakening a type at the end of the context, and then applying another
lifted weakening should be the same as first weakening somewhere and
then weakening at the end. We first assume this equation for types,
then the equation $\q[\gamma^+] = \q$ becomes well-typed (in the
metatheory). The equation for $b[\p][\gamma^+]$ has the same shape as
the newly assumed rule for types. Thus we add the following three
equations.
\begin{alignat*}{10}
& [\p][^+] && : B[\p][\gamma^+] = B[\gamma][\p] \\
& [\p][^+] && : b[\p][\gamma^+] = b[\gamma][\p] \\
& \q[^+] && : \q[\gamma^+] = \q
\end{alignat*}
The second and third equations only make sense because of the first
equation. The phenomenon that later equations (or operations) depend
on previous equations is usual in the world of GATs. In fact, the
later equations cannot even be stated without having some previous
equations.

Now that we have $[\p][^+]$ for types, we can derive the instantiation
rule for nondependent function space as follows.
\begin{alignat*}{10}
  & (A\Ra B)[\gamma] && {=} \\
  & (\Pi\,A\,(B[\p]))[\gamma] && {=}(\Pi[]) \\
  & \Pi\,(A[\gamma])\,(B[\p][\gamma^+])\,\, && {=}([\p][^+]) \\
  & \Pi\,(A[\gamma])\,(B[\gamma][\p]) && {=} \\
  & (A[\gamma])\Ra(B[\gamma])
\end{alignat*}

So far our only terms are variables, but we would like to define
functions via lambda abstraction. Abstraction takes a term in a
context extended by the domain of the function. It comes with a rule
for instantiation analogous to $\Pi[]$.
\begin{alignat*}{10}
& \lam && : \Tm\,(\Gamma\ext A)\,B\ra\Tm\,\Gamma\,(\Pi\,A\,B) \\
& \lam[] && : (\lam\,t)[\gamma] = \lam\,(t[\gamma^+])
\end{alignat*}
Note that the equation $\lam[]$ only makes sense because of the
previous equation $\Pi[]$: the left hand side is in
$\Tm\,\Delta\,(\Pi\,A\,B[\gamma])$, in the right hand side,
$t[\gamma^+] : \Tm\,(\Delta\ext A[\gamma])\,(B[\gamma^+])$, hence the
right hand side is in
$\Tm\,\Delta\,(\Pi\,(A[\gamma])\,(B[\gamma^+]))$.

The functions $\lambda x.x$ and $\lambda x\,y.x$ can be defined in our
syntax as $\lam\,\q : \Tm\,\Gamma\,(A\Ra A)$ and
$\lam\,(\lam\,(\q[\p])) :
\Tm\,\Gamma\,\big(\Pi\,A\,(B\Ra(A[\p]))\big)$ which make sense for all
$\Gamma$, $A$, $B$.

The application operation for dependent function space is a bit
tricky, because the return type depends on the input: the argument of
the function will appear in the return type. We write application by
an infix $\blank\cdot\blank :
\Tm\,\Gamma\,(\Pi\,A\,B)\ra(a:\Tm\,\Gamma\,A)\ra\Tm\,\Gamma\,B'$,
where $B' : \Ty\,\Gamma$ should be $B:\Ty\,(\Gamma\ext A)$ where the
last variable is \emph{substituted} (instantiated) by $a$. For this,
we introduce a new element of $\Sub$ called single substitution which
goes the opposite way of $\p$. Now we can use instantiation
$\blank[\blank]$ to substitute the last variable in $B$.
\begin{alignat*}{10}
& \langle\blank\rangle && : \Tm\,\Gamma\,A\ra\Sub\,\Gamma\,(\Gamma\ext A) \\
& \blank\cdot\blank && : \Tm\,\Gamma\,(\Pi\,A\,B)\ra(a:\Tm\,\Gamma\,A)\ra\Tm\,\Gamma\,(B[\langle a\rangle])
\end{alignat*}
We could have introduced new, separate sorts and $\blank[\blank]$
operations for weakenings and substitutions, but we merge them for
simplicity. There is no need for separation: just as weakenings can be
lifted, single substitutions can also be lifted, and the
weakening-rules $\Pi[]$, $\lam[]$ also work for single
substitutions. We won't have more ways to introduce elements of
$\Sub$, and there are no equations on $\Sub$. An element of $\Sub$ is
either a single weakening $\p$ lifted a couple of times, or a single
substitution lifted a couple of times. We call elements of $\Sub$
substitutions for simplicity.

With the introduction of $\blank\cdot\blank$, we need a new
substitution rule, but again it only makes sense with an extra
equation on types saying that first substituting the last variable and
then an arbitrary substitution is the same as first the lifted version
of the arbitrary substitution that does not touch the last variable,
and then substituting the last variable.
\begin{alignat*}{10}
& [\langle\rangle] && : B[\langle a\rangle][\gamma] = B[\gamma^+][\langle a[\gamma]\rangle] \\
& {\cdot}[] && : (t\cdot a)[\gamma] = (t[\gamma])\cdot(a[\gamma])
\end{alignat*}
Following the introduction of the operation $\langle\blank\rangle$, we
need to explain how it acts on variables. Given a variable in the
middle of the context (a term $b[\p]$), substituting its last variable
simply returns $b$. Substituting the last variable $\q$ reads out the
term from $\langle\blank\rangle$. As usual, the rules only typecheck
if we have an equation for types.
\begin{alignat*}{10}
& [\p][\langle\rangle] && : B[\p][\langle a\rangle] = B \\
& [\p][\langle\rangle] && : b[\p ][\langle a\rangle] = b \\
& \q[\langle\rangle] && : \q[\langle a\rangle] = a
\end{alignat*}
We don't have a separate sort of variables, so the rule
$[\p][\langle\rangle]$ holds not only for variables, but for arbitrary
terms. This is not an issue, as weakening a term at the end of its
context, and then substituting the newly introduced variable is the
same as not doing anything.

We add the computation and uniqueness rule for function space, where
the uniqueness rule again needs an extra equation on the codomain type
of the function to make sense. This is the typal version of the
equation mentioned in the introduction.
\begin{alignat*}{10}
& \Pi\beta && : \lam\,t\cdot a = t[\langle a\rangle] \\
& [\p^+][\langle\q\rangle] && : B[\p^+][\langle\q\rangle] = B \\
& \Pi\eta && : t = \lam\,(t[\p]\cdot\q)
\end{alignat*}

Finally, we add the rules for universes. A universe is a type
containing codes for types, this is witnessed by $\El$ (elements) and
$\cd$ (code), which make an isomorphism between terms of type $\U\,i$
and types of level $i$. All three operations come with substitution
rules ($\El[]$ and $\cd[]$ are interderivable, so it would be enough
to assume one of the two, but we add both to completely align with the
algorithm \cite{DBLP:conf/fscd/KaposiX24} generating the rules from
the SOGAT (\ref{eq:tt})).
\begin{alignat*}{10}
  & \U && : (n:\N)\ra\Ty\,\Gamma\,(1+n) \\
  & \El && : \Tm\,\Gamma\,(\U\,i) \ra \Ty\,\Gamma\,i \\
  & \cd && : \Ty\,\Gamma\,i\ra\Tm\,\Gamma\,(\U\,i) \\
  & \U\beta && : \El\,(\cd\,A) = A \\
  & \U\eta && : \cd\,(\El\,\hat{A}) = \hat{A} \\
  & \U[] && : (\U\,i)[\gamma] = \U\,i \\
  & \El[] && : (\El\,\hat{A})[\gamma] = \El\,(\hat{A}[\gamma]) \\
  & \cd[] && : (\cd\,A)[\gamma] = \cd\,(A[\gamma])
\end{alignat*}

@@
